---
title: "RPI functions"
format: html
editor: visual
---

RPI function: RPI is a quantity used to rank sports teams based upon a team's wins and losses and strength of schedule.

W1 = winning percentage W2 = Opponent's winning percentage W3 = Opponent's Opponent's winning percentage

W1 is calculated by taking a team's wins divided by the number of games it has played

W2 is calculated by taking the average of the W1 for each of the team's opponents with the requirement that all games against the team in question are removed from the equation

W3 is calculated by taking the average of each Opponent's OWP. RPI function for PPG

```{r}
# PPG function takes in a schedule of games
rpi_function_ppg <- function(games, w1=(0.25), w2=(0.5), w3=(0.25)){
  # w1, w2, w3 are weights
  
  stand <- standings2(games)
  
  games <- games %>%
    rename(team = home_team, opponent = away_team, team_score = home_score, opp_score = away_score)
  
  games_2 <- games %>%
    rename(team = opponent, opponent = team, team_score = opp_score, opp_score = team_score)
    
  
  double_table <- bind_rows(games, games_2)
  
  ppg1 <- left_join(double_table, stand, by = c("opponent" = "School")) %>%
    select(team, PPG, 'Win Percentage')%>%
    rename(opp_ppg = PPG)
  
  opp_ppg <- ppg1 %>%
    group_by(team) %>%
    mutate(avg_opp_ppg = mean(opp_ppg)) %>%
    distinct(team, .keep_all = T) %>%
    select(team, avg_opp_ppg)
  
  new_teams_ppg <- left_join(stand, opp_ppg, by = c("School" = "team")) %>%
  rename("team" = "School")

  
  ppg2 <- left_join(double_table, new_teams_ppg, by = c("opponent" =
"team")) %>%
    select(-c(Date, opp_score, team_score)) %>%
    rename(opp_ppg = PPG)
  
    
new_opp_ppg <- ppg2 %>%
    group_by(team) %>%
    mutate(avg_opp_opp_ppg = mean(avg_opp_ppg)) %>%
    distinct(team, .keep_all = T) %>%
    select(team, avg_opp_opp_ppg) 


  
  
  full_teams_ppg <- left_join(new_teams_ppg, new_opp_ppg, by = "team") %>%# to add opponents opponents points per game
    mutate(RPI = (PPG*w1+avg_opp_ppg*w2+avg_opp_opp_ppg*w3)) %>%
    arrange(desc(RPI)) %>%
    select(team, RPI, PPG, avg_opp_ppg, avg_opp_opp_ppg) %>%
    rename('Team' = team, 'RPI' = RPI, 'Points per game (PPG)' = PPG, 'Opponent\'s PPG' = avg_opp_ppg, 'Opponent\'s Opponent\'s PPG' = avg_opp_opp_ppg) %>%
    mutate_if(is.numeric, round, digits = 2)
  
  return(full_teams_ppg)
}
```

Test RPI function

```{r}
test_rpi_wp <- rpi_function_wp(schedule_function(schedule = D3_2023, conf = "LL"))
```

RPI function for Win Percentage:

```{r}
rpi_function_wp <- function(games, w1=(0.25), w2=(0.5), w3=(0.25)){
  
  stand <- standings2(games)
  
  games <- games %>%
    rename(team = home_team, opponent = away_team, team_score = home_score, opp_score = away_score)
  
  games_2 <- games %>%
    rename(team = opponent, opponent = team, team_score = opp_score, opp_score = team_score)
    
  
  double_table <- bind_rows(games, games_2)
  
  wp1 <- left_join(double_table, stand, by = c("opponent" = "School")) %>%
    select(team, PPG, 'Win Percentage')%>%
    rename(opp_wp = 'Win Percentage')
  

  opp_wp <- wp1 %>%
    group_by(team) %>%
    mutate(avg_opp_wp = mean(opp_wp)) %>%
    distinct(team, .keep_all = T) %>%
    select(team, avg_opp_wp)
  
  new_teams_wp <- left_join(stand, opp_wp, by = c("School" = "team")) %>%
  rename("team" = "School")
  
  wp2 <- left_join(double_table, new_teams_wp, by = c("opponent" =
"team")) %>%
    select(-c(Date, opp_score, team_score)) %>%
    rename(opp_wp = 'Win Percentage')
  
  new_opp_wp <- wp2 %>%
    group_by(team) %>%
    mutate(avg_opp_opp_wp = mean(avg_opp_wp)) %>%
    distinct(team, .keep_all = T) %>%
    select(team, avg_opp_opp_wp) 
  
  full_teams_wp <- left_join(new_teams_wp, new_opp_wp, by = "team") %>% # to add opponents opponents WP
    rename(WP = 'Win Percentage') %>%
    mutate(RPI = (WP*w1+avg_opp_wp*w2+avg_opp_opp_wp*w3)) %>%
    arrange(desc(RPI)) %>%
    select(team, RPI, WP, avg_opp_wp, avg_opp_opp_wp) %>%
    rename('Team' = team, 'RPI' = RPI, 'Winning Percentage (WP)' = WP, 'Opponent\'s WP' = avg_opp_wp, 'Opponent\'s Opponent\'s WP' = avg_opp_opp_wp) %>%
    mutate_if(is.numeric, round, digits = 2)
  
  return(full_teams_wp)
}
```
